

%%% Choose a language %%%

\newif\ifEN
\ENtrue   % uncomment this for english
%\ENfalse   % uncomment this for czech

%%% Configuration of the title page %%%

\newif\ifMFF
\MFFtrue % comment this out for the version with a big UK university logo
\def\UKName{Charles University in Prague} %this is only used in UK-logo-version
\def\UKFaculty{Faculty of Mathematics and Physics}

% Thesis type names, as used in several places in the title
\def\ThesisTypeTitle{\ifEN BACHELOR THESIS \else BAKALÁŘSKÁ PRÁCE \fi}
%\def\ThesisTypeTitle{\ifEN MASTER THESIS \else DIPLOMOVÁ PRÁCE \fi}
%\def\ThesisTypeTitle{\ifEN RIGOROUS THESIS \else RIGORÓZNÍ PRÁCE \fi}
%\def\ThesisTypeTitle{\ifEN DOCTORAL THESIS \else DISERTAČNÍ PRÁCE \fi}
\def\ThesisGenitive{\ifEN bachelor \else bakalářské \fi}
%\def\ThesisGenitive{\ifEN master \else diplomové \fi}
%\def\ThesisGenitive{\ifEN rigorous \else rigorózní \fi}
%\def\ThesisGenitive{\ifEN doctoral \else disertační \fi}
\def\ThesisAccusative{\ifEN bachelor \else bakalářskou \fi}
%\def\ThesisAccusative{\ifEN master \else diplomovou \fi}
%\def\ThesisAccusative{\ifEN rigorous \else rigorózní \fi}
%\def\ThesisAccusative{\ifEN doctoral \else disertační \fi}



%%% Fill in your details %%%

% (Note: \xxx is a "ToDo label" which makes the unfilled visible. Remove it.)
\def\ThesisTitle{Modern evolutionary strategies for reinforcement learning problems}
\def\ThesisAuthor{Michal Pospěch}
\def\YearSubmitted{2021}

% department assigned to the thesis
\def\Department{Department of Theoretical Computer Science and Mathematical Logic}
% Is it a department (katedra), or an institute (ústav)?
\def\DeptType{Department}

\def\Supervisor{Mgr. Roman Neruda, CSc.}
\def\SupervisorsDepartment{Department of Theoretical Computer Science and Mathematical Logic}

% Study programme and specialization
\def\StudyProgramme{Computer science}
\def\StudyBranch{General computer science}

\def\Dedication{%
Dedication. \xxx{It is nice to say thanks to supervisors, friends, family, book authors and food providers.}
}

\def\AbstractEN{%
Evolutionary strategies are one of many approaches to solving reinforcement learning tasks. This thesis explores two modern approaches based on them, OpenAI-ES and NS-ES (and its extensions) which utilises novelty search. They are being studied on two environments, Cartpole-swingup and Slimevolley. On Cartpole-swingup they all have some success while performance on Slimevolley is really sensitive to initial seed.
}

\def\AbstractCS{%
Evoluční strategie jsou jeden z přístupů k řešení úloh zpětnovazebného učení. V této práci jsou zkoumány dva moderní přístupy na nich založení, OpenAI-ES a NS-ES (a jeho rozšíření), který využívá hledání novosti. Jsou zkoumány na dvou prostředích, Cartpole-swingup a Slimevolley. Na Cartpole-swingup mají nějaký úspěch všechny, na Slimevolley je chování silně citlivé na úvodní seed.
}

% 3 to 5 keywords (recommended), each enclosed in curly braces.
% Keywords are useful for indexing and searching for the theses by topic.
\def\Keywords{%
{evolutionary strategies} {reinforcement learning} {novelty search} {neuroevolution}
}

% If your abstracts are long and do not fit in the infopage, you can make the
% fonts a bit smaller by this setting. (Also, you should try to compress your abstract more.)
% Alternatively, consider increasing the size of the page by uncommenting the
% geometry modification in thesis.tex.
\def\InfoPageFont{}
%\def\InfoPageFont{\small}  %uncomment to decrease font size

\ifEN\relax\else
% If you are writing a czech thesis, you additionally need to fill in the
% english translation of the metadata here!
\def\ThesisTitleEN{\xxx{Thesis title in English}}
\def\DepartmentEN{\xxx{Name of the department in English}}
\def\DeptTypeEN{\xxx{Department}}
\def\SupervisorsDepartmentEN{\xxx{Superdepartment}}
\def\StudyProgrammeEN{\xxx{study programme}}
\def\StudyBranchEN{\xxx{study branch}}
\def\KeywordsEN{%
\xxx{{key} {words}}
}
\fi
